---
title: "RProgAssingment3"
author: "Thomas J. Keller, PhD"
date: "August 27, 2014"
output: html_document
---

This is the documentation for August 2014 Coursera course 'RProgramming', assingment 3.

## Data source notes   
* Original data at http://hospitalcompare.hhs.gov 
* Data apparently from 12 months prior to July 2012  
* Subsets extracted by Coursera 'R Programming' staff, August 2014  
** https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2FProgAssignment3-data.zip  
```{r}
dir("Assignment3-data/")
```

* The 'Hospital_Revised_Flatfiles.pdf' is the 'code book' for the csv data files.   
* outcome-of-care-measures.csv contains 30-ay mortality and readmission rates for  
** 'Heart.Attack', 'Heart.Failure' and 'Pneumonia'  
* "hospital-data.csv" contains information about the hospitals included in the study

### Background to data  
* including to determine whether hospitals should be fined for the quality of care  
* http://goo.gl/JAXFX  

## Focus of Assigment
* variables for 'Table 19' ("outcome-of-care-measures.csv")  
* variables for 'Table 11' ("hospital-data.csv")  

### Data exploration  
* number of columns  
* number of rows  
* sample the relevant data (columns: 11, 17, 23)  
```{r echo=FALSE}
outcome <- read.csv("Assignment3-data/outcome-of-care-measures.csv", colClasses="character")
ncol(outcome)
nrow(outcome)
colnames(outcome)
# the columns of interest are 2, 7, 11, 17 and 23 for respectively:
## Hospital.Name, State, Heart.Attack mortality, Heart.Failur mortality and Pneumonia mortality
outcome[1:20,c(2, 7, 11,17,23)]
```

#### Replace 'Not Available' values with NA
```{r}
## re-read the csv file, replacing 'Not Available' with NA
outcome <- read.csv("Assignment3-data/outcome-of-care-measures.csv", na.strings = "Not Available")
outcome[1:20,c(2, 7, 11,17,23)]
```

### subset the columns of interest, and rename columns for convenience  
```{r}
mortality <- outcome[, c(2, 7, 11,17,23)]
colnames(mortality) <- c("hospital", "state", "heart attack","heart failure","pneumonia")
str(mortality)
```


* histogram of '30-day death rates'  
```{r echo=FALSE}
hist(as.numeric(outcome[,11]), main="30-day Death Rate for All Hospitals")
hist(as.numeric(mortality$heart_attack))
```


## Finding the best hospital in a state
### final version  
```{r}
best <- function(state, outcome) {
      ## read outcome data in downloaded csv file
      outcome.df <- read.csv("Assignment3-data/outcome-of-care-measures.csv", 
                          colClasses="character",
                          na.strings = "Not Available")
      ## subset columns of interest: hospital, state and three disease outcomes
      mortality <- outcome.df[, c(2, 7, 11,17,23)]
      ## fix numeric values (now columns: 3:5)
      m.df <- sapply(3:5, function(x) as.numeric(mortality[,x]))
      mortality <- cbind(mortality[,1:2],m.df)
      colnames(mortality) <- c("hospital", "state", "heart_attack",
                               "heart_failure","pneumonia")
      ## check that 'state' and 'outcome' arguments are valid; stop() and
      ### return "invalid state", or "invalid outcome" for bad input,
      ### "missing argment" if missing
      if(missing(state)) stop("missing argument")
      if(missing(outcome)) stop("missing argument")
      ## allow variations in case
      state = toupper(state)
      outcome = tolower(outcome)
      ## create vector of valid state names
      states <- as.character(unique(mortality$state))
      if(is.na(states[match(state,states)])) stop("invalid state")
      ## and vector of valid outcome (disease) values
      diseases <- c("heart attack","heart failure","pneumonia")
      if(is.na(diseases[match(outcome,diseases)])) stop("invalid outcome")
      ## sub "_" for space in disease (outcome) name
      disease <- sub("\\s+","_",outcome) 
      ## return hospital in 'state' with lowest 30-day death rate (mortality)
      ### example best("TX", "heart attack") returns [1] "CYPRESS FAIRBANKS..."
      ### hospital names should be sorted alphabetically, return the first for ties
      mystate.df <- mortality[mortality$state == state,]
      stateorder <- order(as.numeric(mystate.df[,which(names(mystate.df) == disease)]),
                          mystate.df[,1])
      besthospital <- as.character(mystate.df[stateorder[1],1])
      return(besthospital)
}
```

## Ranking hospitals by outcome in a state  
### final version  
```{r}
rankhospital <- function(state, outcome, num = "best") {
       ## read outcome data in downloaded csv file
      outcome.df <- read.csv("Assignment3-data/outcome-of-care-measures.csv", 
                          colClasses="character",
                          na.strings = "Not Available")
      ## subset columns of interest: hospital, state and three disease outcomes
      mortality <- outcome.df[, c(2, 7, 11,17,23)]
      ## fix numeric values (now columns: 3:5)
      m.df <- sapply(3:5, function(x) as.numeric(mortality[,x]))
      mortality <- cbind(mortality[,1:2],m.df)

      colnames(mortality) <- c("hospital", "state", "heart_attack",
                               "heart_failure","pneumonia")
      ## check that 'state' and 'outcome' arguments are valid; stop() and
      ### return "invalid state", or "invalid outcome" for bad input,
      ### "missing argment" if missing
      if(missing(state)) stop("missing argument")
      if(missing(outcome)) stop("missing argument")
      ## allow variations in case
      state = toupper(state)
      outcome = tolower(outcome)
      ## create vector of valid state names
      states <- as.character(unique(mortality$state))
      if(is.na(states[match(state,states)])) stop("invalid state")
      ## and vector of valid outcome (disease) values
      diseases <- c("heart attack","heart failure","pneumonia")
      if(is.na(diseases[match(outcome,diseases)])) stop("invalid outcome")
      ## sub "_" for space in disease (outcome) name
      disease <- sub("\\s+","_",outcome)
      if(is.na(num)) stop("invalid num")
      ## return hospital in 'state' with that rank for 30-day death rate (mortality)
      ### a num (rank) that is greater than the largest number for rank should return NA
      mystate.df <- mortality[mortality$state == state,]
      stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],
                          mystate.df[,1])
        ## recode for "best"  and "wworst" 
      if(is.numeric(num)) rownumber <- stateorder[num]
      if(!is.numeric(num)) {
      num <- if(num == "best", 1)
      num <- ifelse(num == "worst" && disease == "heart_attack", 
        which(mystate.df$heart_attack == max(mystate.df$heart_attack, na.rm=T)), num)
      num <- ifelse(num == "worst" && disease == "heart_failure",
        which(mystate.df$heart_failure == max(mystate.df$heart_failure, na.rm=T)), num)
      
      rownumber <- num
      }
      ## use num to retrieve ranked hospital
      rankedhospital <- as.character(mystate.df[rownumber,1])
      return(rankedhospital)
}
```

## Ranking hospitals in all states
```{r}
rankall <- function(outcome, num = "best") {
        require(dplyr)
        ## read outcome data in downloaded csv file
        outcome.df <- read.csv("Assignment3-data/outcome-of-care-measures.csv",
                               colClasses="character",
                               na.strings = "Not Available")
        ## subset columns of interest: hospital, state and three disease outcomes
        mortality <- outcome.df[, c(2, 7, 11,17,23)]      
        ## fix numeric values (now columns: 3:5)
        m.df <- sapply(3:5, function(x) as.numeric(mortality[,x]))
        mortality <- cbind(mortality[,1:2],m.df)
        
        colnames(mortality) <- c("hospital", "state", "heart_attack",
                                 "heart_failure","pneumonia")
        ## check the 'outcome' arguments is valid; stop() and
        ### return "invalid outcome" for bad input,
        ### "missing argment" if missing argument)
        if(missing(outcome)) stop("missing argument")
        outcome = tolower(outcome)
        diseases <- c("heart attack","heart failure","pneumonia")
        if(is.na(diseases[match(outcome,diseases)])) stop("invalid outcome")
        ## sub "_" for space in disease (outcome) name
        disease <- sub("\\s+","_",outcome)
        ## create a vector of valid state names
        states <- as.character(sort(unique(mortality$state)))
        mort_df <- tbl_df(mortality)
        # capture column for outcome argument
        disease.cols <- as.vector(sapply(colnames(mort_df), function(x) sub("\\s+","_",x)))
        col_disease <- match(disease,disease.cols)
        states.list <- lapply(states, function(x) mort_df[mort_df$state == x,c(1,2,col_disease)])
        ## fix num for "best" 
        ifelse(num == "best", num <- 1, num)
        ## create list of states with outcome rank = num
        output.list <- lapply( 1:length(states), function(x,y,z) FUN={
                mystate_df <- states.list[[x]]
                colnames(mystate_df) <- c("hospital","state","disease")
                filter_s_df <- filter(mystate_df,!is.na(disease))
                sorted_s_df = arrange(filter_s_df,disease,hospital)
                ifelse(num == "worst", num <- length(sorted_s_df$disease), num)
                sorted_s_df[num,1:2]   
        }, y=disease, z=num)
        
        output.df <- rbind_all(output.list)

}
```

### functional units
* read outcome and subset relevant columns, replace 'Not Available' with NA
* check arguments
** 'state' must be a member of the mortality$state list of unique values
** check 'outcome' arg is c("heart attack", "heart failure", "pneumonia")  
```{r}
states <- as.character(unique(mortality$state))
diseases <- c("heart attack", "heart failure", "pneumonia")
check <- function(state, outcome) {
 if(missing(state)) stop("missing argument")
 if(missing(outcome)) stop("missing argument")
 ## allow variations in case
 state = toupper(state)
 outcome = tolower(outcome)
 states <- as.character(unique(mortality$state))
 if(is.na(states[match(state,states)])) stop("invalid state")
 if(is.na(diseases[match(outcome,diseases)])) stop("invalid outcome")
 args <- paste(outcome,state, sep= " in ")
 return(args)
}

```

* recode num
```{r}
if(!is.numeric(num)) {
                        if(num == "best") num <- 1
                        if(num == "worst") 
                                num <- which(mystate.df[,3] == max(mystate.df[,3],
                                                                   na.rm=T))
                        }
```


* subset by state
* best outcome  
** be sure to include validation in the final function  
```{r}
stateoutcome <- function(state, outcome) {
      ## fix outcome names
      disease <- sub("\\s+","_",outcome)          
      ## fix column names
      colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
      mystate.df <- mortality[mortality$state == state,]
      stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],mystate.df[,1])
      besthospital <- as.character(mystate.df[stateorder[1],1])
      return(besthospital)
}
```

* subset by state
* rank the outcome  
** be sure to include validation in the final function  
```{r}
rankhospital <- function(state, outcome, num = "best"){
      ## validate arguments
      # num has a default, so no call to missing(num)
      ## fix outcome names
      disease <- sub("\\s+","_",outcome)   
      ## fix column names
      colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
      mystate.df <- mortality[mortality$state == state, c(1,which(names(mortality) == disease))]
      #stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],mystate.df[,1], na.last = T) make values numeric!!
      ## recode "best" to 1 and "wworst" to length(stateorder)
      num <- ifelse(num == "best", 1, num)
      num <- ifelse(num == "worst", length(stateorder), num)
      num <- as.numeric(num)
      if(is.na(num)) stop("invalid num")
      ## order by outcome and hospital name
      mystate.df <- mystate.df[stateorder[num],c(1,which(names(mystate.df) == disease))]
      return(mystate.df)
}
```

* sortorder with options for 'num' argument, no state specified
```{r}
orderindex <- function(x) {
        mort_df <- tbl_df(mortality)
        # for apply: x is index to states, y is outcome (disease)
        outcome <- "heart_attack"
        states_comp.df <- sapply(1:length(states), function(x,y) {
                mystate.df <- filter(mort_df,state == states[x])
                ok <- complete.cases(mystate.df[,c(1,2,mystate.df$y)])
                mystate.df[unlist(ok),c(1,2,mystate.df$y)]
               }, y = outcome
                )

}
        ## recode for "best"  and "wworst" 
      if(is.numeric(num)) rownumber <- stateorder[num]
      if(!is.numeric(num)) {
      num <- ifelse(num == "best", 1, num)
      num <- ifelse(num == "worst" && disease == "heart_attack", 
        which(mystate.df$heart_attack == max(mystate.df$heart_attack, na.rm=T)), num)
      num <- ifelse(num == "worst" && disease == "heart_failure",
        which(mystate.df$heart_failure == max(mystate.df$heart_failure, na.rm=T)), num)
      num <- ifelse(num == "worst" && disease == "pneumonia", 
        which(mystate.df$pneumonia == max(mystate.df$pneumonia, na.rm=T)), num)
      rownumber <- num
      }
```

* apply to all states
* rank outcome and order by outcome and hospital name
** return hospital name and state
```{r}
rankall <- function(disease, num = "best") {
        #install.packages("dplyr")
        require(dplyr)
        # read and fix data with proper columnn names
        colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
        states <- as.character(unique(mortality$state))
        ## subset df for the outcome (disease)
        d.df <- mortality[, c(1,2,match(disease,colnames(mortality)))]
        ## apply to all states, return df of hospital with num (rank) 
        ### for all states
        do.call(rbind, 
                lapply(states, function(x,y) {
                state.df = d.df[d.df$state == x, ]
                sortorder = order(state.df[,3,1])
                state.df[sortorder[y],1:2]
                }, y = 1)
                )

}
```

